---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(doFuture.rng.onMisuse = "ignore")
```

# xiacf: Chatterjee's Rank Correlation for Time Series Analysis

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/xiacf)](https://CRAN.R-project.org/package=xiacf)
[![R-CMD-check](https://github.com/yetanothersu/xiacf/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yetanothersu/xiacf/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.19247735.svg)](https://doi.org/10.5281/zenodo.19247735)
<!-- badges: end -->

The **xiacf** package provides a robust framework for detecting complex non-linear and functional dependence in time series data. Traditional linear metrics, such as the standard Autocorrelation Function (ACF) and Cross-Correlation Function (CCF), often fail to detect symmetrical or purely non-linear relationships. 

This package overcomes these limitations by utilizing **Chatterjee's Rank Correlation ($\xi$)**, offering both univariate ($\xi$-ACF) and multivariate ($\xi$-CCF) analysis tools. It features rigorous statistical hypothesis testing powered by advanced surrogate data generation algorithms (IAAFT and MIAAFT), all implemented in high-performance C++ using `RcppArmadillo`.

## Key Features

* **Non-linear Autocorrelation ($\xi$-ACF):** Detect time-dependent structures that standard linear ACF completely misses (e.g., chaotic systems, volatility clustering).
* **Multivariate Cross-Correlation ($\xi$-CCF):** Uncover hidden non-linear lead-lag relationships between two different time series.
* **MIAAFT Surrogate Testing:** Rigorous null hypothesis testing using Multivariate Iterative Amplitude Adjusted Fourier Transform (MIAAFT). It preserves the exact marginal distributions and the instantaneous (lag-0) cross-correlation while destroying lagged non-linear dependence.
* **Rolling Window Analysis:** Track how non-linear dependencies evolve over time (detecting structural breaks or market regime shifts) with robust parallel processing support via the `future` ecosystem.
* **High Performance:** Core algorithms are heavily optimized in C++ to handle the computationally intensive surrogate iterations.

## Installation

You can install the development version of xiacf from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("yetanothersu/xiacf")
```

*(Note: CRAN submission is currently pending. Once accepted, you can install it via `install.packages("xiacf")`)*

## Quick Start

Here is a basic example showing how to compute and visualize the $\xi$-ACF against a standard linear ACF.

```{r xi-acf-test}
#| fig.cap: "Comparison between standard linear ACF and Chatterjee's Xi-ACF."
#| fig.alt: "A correlogram comparing linear and non-linear dependence."
library(xiacf)
library(ggplot2)

# Generate a chaotic Logistic Map: x_{t+1} = r * x_t * (1 - x_t)
set.seed(42)
n <- 500
x <- numeric(n)
x[1] <- 0.1 # Initial condition
r <- 4.0 # Fully chaotic regime

for (t in 1:(n - 1)) {
  x[t + 1] <- r * x[t] * (1 - x[t])
}

# 1. Run the Xi-ACF test
# Computes up to 20 lags with 100 IAAFT surrogates for significance testing
results <- xi_test(x, max_lag = 20, n_surr = 100)

# Print summary
print(results)

# 2. Visualize the results
# The autoplot method automatically generates a ggplot2 object
autoplot(results)
```

## Multivariate Cross-Correlation ($\xi$-CCF)

The package supports multivariate analysis to detect non-linear lead-lag relationships between two time series. It uses MIAAFT surrogates to preserve both the individual distributions and the lag-0 cross-correlation, providing a rigorous threshold for significance.

The following example demonstrates a scenario where the standard CCF completely fails to detect the relationship, but $\xi$-CCF correctly identifies the lead-lag structure.

```{r xi-ccf-test}
#| fig.cap: "Multivariate Xi-CCF detecting a purely non-linear lead-lag relationship."
#| fig.alt: "A cross-correlogram showing a peak at lag 3 for Xi, while standard CCF remains within the noise bounds."

# Generate a pure non-linear lead-lag relationship
# Y is driven by the absolute value of X from 3 periods ago.
set.seed(42)
n <- 300
# A uniform distribution centered at 0 ensures the linear cross-correlation is zero
X <- runif(n, min = -2, max = 2)
Y <- numeric(n)

for (t in 4:n) {
  Y[t] <- abs(X[t - 3]) + rnorm(1, sd = 0.2)
}

# Run the multivariate Xi-CCF test
ccf_results <- xi_ccf(x = X, y = Y, max_lag = 10, n_surr = 100)

# Visualize the results
# Standard CCF misses the V-shaped relationship, but Xi-CCF correctly detects that X leads Y by 3 periods.
library(ggplot2)
autoplot(ccf_results) +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = "Lag (X leads Y)")
```

## Rolling Window Analysis

For advanced market microstructure or structural break detection, you can run a rolling $\xi$-ACF analysis. This function supports parallel processing using the `future` ecosystem.

```{r xi-acf-rolling}
# Run rolling analysis with a window size of 100 and step size of 10
rolling_res <- run_rolling_xi_analysis(
  x = x,
  window_size = 100,
  step_size = 10,
  max_lag = 5,
  n_surr = 50,
  n_cores = 2 # Set to NULL for sequential execution
)

head(rolling_res)
```

## References

* Chatterjee, S. (2021). A new coefficient of correlation. *Journal of the American Statistical Association*, 116(536), 2009-2022.

## License

This project is licensed under the MIT License - see the LICENSE file for details.
