---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(doFuture.rng.onMisuse = "ignore")
```

# xiacf: Chatterjee's Rank Correlation for Time Series Analysis

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/xiacf)](https://CRAN.R-project.org/package=xiacf)
[![R-CMD-check](https://github.com/yetanothersu/xiacf/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yetanothersu/xiacf/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->
The xiacf package provides a robust framework for detecting complex non-linear and functional dependence in time series data. Traditional linear metrics, such as the standard Autocorrelation Function (ACF), often conflate stochastic noise with structural decay. This package overcomes these limitations by integrating Chatterjee's rank correlation coefficient ($\xi$) with high-performance C++ backend computations. Traditional linear metrics, such as the standard Autocorrelation Function (ACF), often conflate stochastic noise with structural decay. This package overcomes these limitations by integrating **Chatterjee's rank correlation coefficient ($\xi$)** with high-performance C++ backend computations.

## Key Features

* **Fast Computation**: Core engine is written in C++ (`Rcpp` and `RcppArmadillo`) for lightning-fast calculations, even on high-frequency financial data.
* **$\xi$-ACF Test**: Computes Chatterjee's $\xi$ alongside the standard linear ACF for multiple lags (`xi_test()`).
* **Robust Significance Testing**: Automatically generates Iterative Amplitude Adjusted Fourier Transform (IAAFT) surrogate datasets to establish 95% confidence thresholds for non-linear dependence.
* **Rolling Analysis**: Fully parallelized rolling window analysis (`run_rolling_xi_analysis()`) to monitor structural changes over time.
* **Beautiful Visualizations**: Built-in `ggplot2` integration (`autoplot()`) for publication-ready correlograms.

## Installation

You can install the development version of xiacf from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("yetanothersu/xiacf")
```

*(Note: CRAN submission is currently pending. Once accepted, you can install it via `install.packages("xiacf")`)*

## Quick Start

Here is a basic example showing how to compute and visualize the $\xi$-ACF against a standard linear ACF.

```{r xi-acf-test}
#| fig.cap: "Comparison between standard linear ACF and Chatterjee's Xi-ACF."
#| fig.alt: "A correlogram comparing linear and non-linear dependence."
library(xiacf)
library(ggplot2)

# Generate a chaotic Logistic Map: x_{t+1} = r * x_t * (1 - x_t)
set.seed(42)
n <- 500
x <- numeric(n)
x[1] <- 0.1 # Initial condition
r <- 4.0 # Fully chaotic regime

for (t in 1:(n - 1)) {
  x[t + 1] <- r * x[t] * (1 - x[t])
}

# 1. Run the Xi-ACF test
# Computes up to 20 lags with 100 IAAFT surrogates for significance testing
results <- xi_test(x, max_lag = 20, n_surr = 100)

# Print summary
print(results)

# 2. Visualize the results
# The autoplot method automatically generates a ggplot2 object
autoplot(results)
```

## Rolling Window Analysis

For advanced market microstructure or structural break detection, you can run a rolling $\xi$-ACF analysis. This function supports parallel processing using the `future` ecosystem.

```{r xi-acf-rolling}
# Run rolling analysis with a window size of 100 and step size of 10
rolling_res <- run_rolling_xi_analysis(
  x = x,
  window_size = 100,
  step_size = 10,
  max_lag = 5,
  n_surr = 50,
  n_cores = 2 # Set to NULL for sequential execution
)

head(rolling_res)
```

## References

* Chatterjee, S. (2021). A new coefficient of correlation. *Journal of the American Statistical Association*, 116(536), 2009-2022.

## License

This project is licensed under the MIT License - see the LICENSE file for details.
