---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(doFuture.rng.onMisuse = "ignore")
Sys.setlocale("LC_TIME", "C")
```

# xiacf: Chatterjee's Rank Correlation for Time Series Analysis

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/xiacf)](https://CRAN.R-project.org/package=xiacf)
[![R-CMD-check](https://github.com/yetanothersu/xiacf/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yetanothersu/xiacf/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.19247735.svg)](https://doi.org/10.5281/zenodo.19247735)
<!-- badges: end -->

The **xiacf** package provides a robust framework for detecting complex non-linear and functional dependence in time series data. Traditional linear metrics, such as the standard Autocorrelation Function (ACF) and Cross-Correlation Function (CCF), often fail to detect symmetrical or purely non-linear relationships. 

This package overcomes these limitations by utilizing **Chatterjee's Rank Correlation ($\xi$)**, offering both univariate ($\xi$-ACF) and multivariate ($\xi$-CCF) analysis tools. It features rigorous statistical hypothesis testing powered by advanced surrogate data generation algorithms (IAAFT and MIAAFT), all implemented in high-performance C++ using `RcppArmadillo`.

## Key Features

* **Non-linear Autocorrelation ($\xi$-ACF):** Detect time-dependent structures that standard linear ACF completely misses (e.g., chaotic systems, volatility clustering).
* **Multivariate Cross-Correlation ($\xi$-CCF):** Uncover hidden non-linear lead-lag relationships between two different time series.
* **MIAAFT Surrogate Testing:** Rigorous null hypothesis testing using Multivariate Iterative Amplitude Adjusted Fourier Transform (MIAAFT). It preserves the exact marginal distributions and the instantaneous (lag-0) cross-correlation while destroying lagged non-linear dependence.
* **Rolling Window Analysis:** Track how non-linear dependencies evolve over time (detecting structural breaks or market regime shifts) with robust parallel processing support via the `future` ecosystem.
* **High Performance:** Core algorithms are heavily optimized in C++ to handle the computationally intensive surrogate iterations.

## Installation

You can install the development version of xiacf from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("yetanothersu/xiacf")
```

*(Note: CRAN submission is currently pending. Once accepted, you can install it via `install.packages("xiacf")`)*

## Quick Start

Here is a basic example showing how to compute and visualize the $\xi$-ACF against a standard linear ACF.

```{r xi-acf-test}
#| fig.cap: "Comparison between standard linear ACF and Chatterjee's Xi-ACF."
#| fig.alt: "A correlogram comparing linear and non-linear dependence."
library(xiacf)
library(ggplot2)

# Generate a chaotic Logistic Map: x_{t+1} = r * x_t * (1 - x_t)
set.seed(42)
n <- 500
x <- numeric(n)
x[1] <- 0.1 # Initial condition
r <- 4.0 # Fully chaotic regime

for (t in 1:(n - 1)) {
  x[t + 1] <- r * x[t] * (1 - x[t])
}

# 1. Run the Xi-ACF test
# Computes up to 20 lags with 100 IAAFT surrogates for significance testing
results <- xi_test(x, max_lag = 20, n_surr = 100)

# Print summary
print(results)

# 2. Visualize the results
# The autoplot method automatically generates a ggplot2 object
autoplot(results)
```

## Bidirectional $\xi$-CCF Test (Directional Lead-Lag Analysis)

While the standard CCF is symmetric in its linear evaluation, `xi_ccf()` evaluates the **directional** non-linear lead-lag relationship. By default (`bidirectional = TRUE`), it computes both "$X$ leads $Y$" and "$Y$ leads $X$" simultaneously. Thanks to our optimized C++ engine, the reverse direction is computed essentially at zero additional cost by reusing the MIAAFT surrogates.

```{r ccf-example, fig.alt="Directional Xi-CCF correlogram showing a peak at lag 3 in the X leads Y panel, while standard CCF remains within the noise bounds."}
# Generate a pure non-linear lead-lag relationship
# Y is driven by the absolute value of X from 3 periods ago.
set.seed(42)
n <- 300
# A uniform distribution centered at 0 ensures the linear cross-correlation is zero
X <- runif(n, min = -2, max = 2)
Y <- numeric(n)

for (t in 4:n) {
  Y[t] <- abs(X[t - 3]) + rnorm(1, sd = 0.2)
}

# Run the bidirectional Xi-CCF test
ccf_results <- xi_ccf(x = X, y = Y, max_lag = 10, n_surr = 100)

# Visualize the results
# The new autoplot generates a beautiful 2-panel graph showing directional dependence.
# Standard CCF misses the V-shaped relationship, but Xi-CCF correctly detects that X leads Y by 3 periods.

autoplot(ccf_results)
```

## Rolling Window Analysis

For advanced market microstructure or structural break detection, you can run rolling $\xi$-ACF or $\xi$-CCF analyses. These functions support robust parallel processing via the `future` ecosystem and seamlessly integrate with timestamps for intuitive visualization.

```{r xi-ccf-rolling, fig.alt="Heatmap of rolling bidirectional Xi-CCF showing a time-varying non-linear lead-lag relationship."}
library(ggplot2)

# Generate dummy time series data with a structural break
set.seed(123)
dates <- seq(as.Date("2020-01-01"), by = "1 day", length.out = 300)
X <- rnorm(300)
Y <- numeric(300)

# First half (Day 1-150): X leads Y by 3 days (non-linear relationship)
Y[1:150] <- c(rnorm(3), abs(X[1:147])) + rnorm(150, sd = 0.1)
# Second half (Day 151-300): The relationship breaks down (pure noise)
Y[151:300] <- rnorm(150)

# Run rolling bidirectional Xi-CCF with time_index
rolling_res <- run_rolling_xi_ccf(
  x = X,
  y = Y,
  time_index = dates, # Pass the dates directly!
  window_size = 100,
  step_size = 5,
  max_lag = 5,
  n_surr = 50,
  n_cores = 2 # Set to NULL for sequential execution
)

# Visualize the dynamic relationship as a beautiful heatmap
ggplot(rolling_res, aes(x = Window_End_Time, y = Lag, fill = Xi_Excess)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "firebrick", mid = "white", midpoint = 0) +
  facet_wrap(~Direction, ncol = 1) +
  scale_x_date(date_labels = "%Y-%m") +
  labs(
    title = "Rolling Bidirectional Xi-CCF Heatmap",
    subtitle = "Detecting structural breaks in non-linear lead-lag dynamics",
    x = "Date",
    fill = "Excess Xi"
  ) +
  theme_minimal()
```

## References

* Chatterjee, S. (2021). A new coefficient of correlation. *Journal of the American Statistical Association*, 116(536), 2009-2022.

## License

This project is licensed under the MIT License - see the LICENSE file for details.
